### 2024-11-15 teaching/math240f24/draft.tex
lectures this week

\bigskip

@@

\vskip60pt

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}

@@

\proclaim Proposition \advthm. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |F|$.

\proof We shall construct a function $f:E\to F$. Fix one specific $x\in X$;
we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
$$f(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
Note that since $|A|$ is even for all $A\in E$, the cardinality of $f(A)$ is odd (in the first
case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
do by showing that $f$ is an inverse (as an exercise, you might instead try to prove bijectivity from
the definitions of injective and surjective).

We define $g : F\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug

@@

If $a$ is not divisible by $b$, then consider the numbers
$$ \ldots, a - 3b, a-2b, a-b, a, a+b, a+2b, a+3b, \ldots.$$
Let $S$ be the set of these integers that are positive. Symbolically, we have
$$ S = \{a-kb : k\in \ZZ\ \hbox{and}\ a-kb\ge 0\}.$$
By the Archimedean property, there is some $n\in \NN$ such that $n> -a$, which
implies that $nb \ge n > -a$ (here we use the fact that if $b>0$ and $b$ is an integer, then
$b\ge 1$). From this we derive $a > -nb$, and hence $a+nb = a - (-n)b > 0$. This shows
that $S$ is nonempty.

@@

\proclaim\datestamp{08}{xi} Lemma \advthm.
A graph $G=(V,E)$ is bipartite if and only if all of its connected components
are bipartite.

@@

@@

\parenproclaim Lemma {\advthm} (Detour lemma). Let $G$ be a connected graph and let
$\sigma$ be a cycle in $G$. If $G'$ is the graph obtained by deleting one edge
of $\sigma$, then $G'$ is also connected.

\proclaim\datestamp{20}{} Theorem \advthm. A graph $G = (V,E)$ is a tree if and only if
$|E| = |V|-1$ and $G$ is connected.

\proof We prove the forward implication by induction on $n=|V|$, the number of vertices.
Concretely, the statement we shall prove is, {\sl For all $n\ge 1$, for all $G=(V,E)$ with
$|V| = n$, if $G$ is a tree then $|E| = |V|-1$ and $G$ is connected.}

The base case is $n=1$. Then the only possible graph $G$ is simply $\bullet$, a single vertex.
This is a tree since it is connected and contains no cycles. We have $|E| = 0 = 1-1 = |V|-1$.

For the inductive step, let $n\ge 1$ and assume the statement holds for $n$. Let $G = (V,E)$
be a graph with $|V| = n+1$, and assume that $G$ is a tree. Since $n\ge 1$,
$|V| \ge 2$, so there are at least two vertices, and connectedness of $G$ means that there
is at least one edge in $G$. Hence $G$ is a forest with at least one edge, and by
Proposition~{\proptwoleaves}, there are at least two leaves in $G$. Pick one of these leaves
and call it $u$. Form a new graph $G' = (V', E')$ by removing $u$ and the single edge
incident on it; we have

>>>>

@@

\vskip48pt

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}

@@

\proclaim Proposition \advthm. Let $X$ be a finite nonempty set.
Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $D$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |D|$.

\proof We shall construct a function $f:E\to D$. Fix one specific $x\in X$;
we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
$$f(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
Note that since $|A|$ is even for all $A\in E$, the cardinality of $f(A)$ is odd (in the first
case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
with codomain $D$. Now to prove $|E| = |D|$ we will show that $f$ is bijective, which we
do by showing that $f$ is an inverse (as an exercise, you might instead try to prove bijectivity from
the definitions of injective and surjective).

We define $g : D\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $D$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug

@@

If $a$ is not divisible by $b$, then consider the numbers
$$ \ldots, a - 3b, a-2b, a-b, a, a+b, a+2b, a+3b, \ldots.$$
Let $S$ be the set of these integers that are positive. Symbolically, we have
$$ S = \{a-kb : k\in \ZZ\ \hbox{and}\ a-kb\ge 0\}\phantom{D}.$$
By the Archimedean property, there is some $n\in \NN$ such that $n> -a$, which
implies that $nb \ge n > -a$ (here we use the fact that if $b>0$ and $b$ is an integer, then
$b\ge 1$). From this we derive $a > -nb$, and hence $a+nb = a - (-n)b > 0$. This shows
that $S$ is nonempty.

@@

\proclaim\datestamp{07}{xi} Lemma \advthm.
A graph $G=(V,E)$ is bipartite if and only if all of its connected components
are bipartite.

@@

\parenproclaim\datestamp{12}{xi}
Lemma {\advthm} (Detour lemma). Let $G$ be a connected graph and let
$\sigma$ be a cycle in $G$. If $G'$ is the graph obtained by deleting one edge
of $\sigma$, then $G'$ is also connected.

\proclaim Theorem \advthm. A graph $G = (V,E)$ is a tree if and only if
$|E| = |V|-1$ and $G$ is connected.

\proof We prove the forward implication by induction on $n=|V|$, the number of vertices.
Concretely, the statement we shall prove is, {\sl For all $n\ge 1$, for all $G=(V,E)$ with
$|V| = n$, if $G$ is a tree then $|E| = |V|-1$ and $G$ is connected.}
The base case is $n=1$. Then the only possible graph $G$ is simply $\bullet$, a single vertex.
This is a tree since it is connected and contains no cycles. We have $|E| = 0 = 1-1 = |V|-1$.
For the inductive step, let $n\ge 1$ and assume the statement holds for $n$. Let $G = (V,E)$
be a graph with $|V| = n+1$, and assume that $G$ is a tree. Since $n\ge 1$,
$|V| \ge 2$, so there are at least two vertices, and connectedness of $G$ means that there
is at least one edge in $G$. Hence $G$ is a forest with at least one edge, and by
Proposition~{\proptwoleaves}, there are at least two leaves in $G$. Pick one of these leaves
and call it $u$. Form a new graph $G' = (V', E')$ by removing $u$ and the single edge
incident on it; we have $|V'| = n$ and $|E'| = |E|-1$. The graph $G'$ is a tree, because
for each $v,w\in V'$, there is a unique path from $v$ to $w$ in $G$, and this path
does not pass through $u$ (since $u$ is a leaf). Hence
$$ |E| = |E'| + 1  = |V'| - 1 + 1 = |V|-1,$$
where in the second equality we have used the induction hypothesis.

Now assume that $|E| = |V|-1$ and $G$ is connected. We want to show that $G$ has no cycles.
For a contradiction, suppose it does; call it $\sigma$. Remove an edge from $\sigma$. By
the detour lemma, this resulting graph is still connected. If this graph still has a cycle,
remove an edge from it again, and repeat this process until no cycles remain. This graph
$G'$ still has $V$ as its vertex set, but it has a new edge set $E'$ with $|E'|< |E|$.
But $G'$ is now a tree, since it contains no cycles but is still connected, so by
the previous paragraph, we have $|E'| = |V|-1 = |E|$. The contradiction completes the proof.\slug

We are now able to answer the following extremal question: {\sl How many edges can a graph
on $n$ vertices have if it does not contain a cycle (i.e., if it is a forest?}

\proclaim Corollary \advthm. Let $F = (V,E)$ be a forest. Then $|E| \le |V|-1$.

\proof Let $k\ge 1$ denote the number of connected components in $F$, and number the
connected components $C_1, C_2,\ldots,C_k$.
Join $C_1$ to $C_2$ by an edge, $C_2$ to $C_3$ by an
edge, and so on. This creates a connected graph $G = (V, E')$ with $|E'| = |E| + k-1$, and $G$ must
be a tree, since adding these edges does not introduce a cycle. By the previous
theorem, $|E'| = |V|-1$, so $|E| \le |V|-1$.\slug

\advsect Eulerian trails and circuits

Recall that a walk that does not repeat any vertex is called a path. A similar definition
is that of a {\it trail}: this is a walk that does not repeat any {\it edge}.
A walk is called an {\it Eulerian trail} if it uses every edge of $G$ exactly once, and an
{\it Eulerian circuit} if it is an Eulerian trail and it is closed (its endpoints are
the same vertex).

The term ``Eulerian'' refers to L.~Euler, who in 1735 answered the following question:
{\sl Is it possible to cross every bridge in K\"onigsberg (a city in Prussia),
without having to repeat any bridges?} The situation is illustrated in Fig.~6.
\midinsert
$$\epsfxsize=220pt\epsfbox{konigsberg.ps}$$
\vskip5pt
\caption{The seven bridges of K\"onigsberg in 1735.}
\endinsert
\goodbreak
To represent this map as a graph $G = (V,E)$, we will have to allow multiple edges between vertices.
There are a couple ways to formalise this set-theoretically, but we will dispense with the details
here and simply think of $E$ as a special type of set
that allows for copies of the same edge to be included multiple times. The graph $G$ representing
K\"onigsberg is depicted in Fig.~7.
\midinsert
$$\epsfbox{konigsberggraph.ps}$$
\vskip5pt
\caption{The graph representing the seven bridges of K\"onigsberg.}
\endinsert
\goodbreak
What Euler actually proves in his paper (which did not appear in print until 1741) is that
it is impossible to cross every bridge in K\"onigsberg exactly once; i.e., $G$ does not
contain an Eulerian trail.
First we characterise the graphs that contain Eulerian circuits, as we'll need this to later
characterise graphs with Eulerian trails.

\proclaim Theorem \advthm. Let $G = (V,E)$ be a connected graph with a finite number of vertices
and edges, where multiple edges between the same two vertices are allowed. Then
$G$ has an Eulerian circuit if and only if all vertices have even degree.

\proof Suppose $G$ has an Eulerian circuit $\sigma = (v_0, v_1, \ldots, v_n)$ where
$v_0 = v_n$. For each $v_i\ne v_0$, each visit uses two edges, one edge to enter $v_i$ and
one to exit it, so $\deg(v_i)$ is even for all $0\le i < n$. For $v_0 = v_n$, there is one
edge that leaves it at the beginning, two edges for each visit during the circuit, and then
one returning to it at the end, so $\deg(v_0)$ is even as well.

We prove the reverse implication by induction on $m = |E|$. That is, we prove
that for all integers $m\ge 0$, if $G = (V,E)$ with $|E| = m$ and all vertices in $G$ have
even degree, then $G$ has an Eulerian circuit.

For the base case $m=0$, the stipulation
that $G$ be connected forces
$G = \bullet$, and this graph has an Eulerian circuit (the trivial walk).

Now let $m\ge 0$ and suppose the statement is true for all integers less than or equal to $m$.
Let $G = (V,E)$ with $|E| = m+1$. Since there is at least one edge in $G$ now, $G$ cannot be
a tree. This is because every tree with at least one edge contains at least two leaves,
and leaves have (odd) degree $1$. Let $\sigma$ be a cycle in $G$ and form $G' = (V,E')$
be removing every edge in $\sigma$. This graph may no longer be connected, but since we have removed
a cycle, every vertex in $G'$ either has the same degree as in $G$ (if the vertex
was not in $\sigma$), or degree two less than in $G$ (if the vertex was in $\sigma$). So all
vertices in $G'$ have even degree. Let $k$ be the number of connected components in
$G'$; number them $H_1,\ldots,H_k$. Each has at most $m$ edges, is connected, and all vertices
have even degree, so each $H_i$ contains an Eulerian circuit, by induction. We build an Eulerian
circuit in $G$ as follows.
\medskip
\item{i)} Start at any vertex of $\sigma$.
\smallskip
\item{ii)} Follow $\sigma$.
\smallskip
\item{iii)} Each time we reach a vertex of a connected component $H_i$ of $G'$ for the first
time, follow its Eulerian circuit.
\smallskip
\item{iv)} Once we return to the beginning of this circuit, continue along $\sigma$.
\smallskip
\item{v)} Repeat this process until we return to the vertex of $\sigma$ that we started at.
\medskip
This gives an Eulerian circuit of $G$.\slug

Now we characterise Eulerian trails.

\proclaim Theorem \advthm. Let $G = (V,E)$ be a graph (with multiple edges allowed). Then $G$ contains
an Eulerian trail that isn't an Eulerian circuit
if and only if exactly two vertices of $G$ have odd degree.

\proof Suppose that $G$ contains an Eulerian trail $\sigma = (v_0,v_1,\ldots,v_n)$
in which $v_0\ne v_n$. By a similar reasoning to the first paragraph of the previous proof,
every vertex in the walk has even degree, except both $v_0$ and $v_n$ must have odd degree,
since starting at $v_0$ we must leave for the first time via one edge, and at the end we enter $v_n$
one last time without exiting.

Now assume that $G$ contains exactly two odd-degree vertices. By the previous theorem, there does
not exist an Eulerian circuit in $G$. Let $u$ and $v$ be the two odd-degree vertices.
Add a new edge $e = uv$, even if there was already an edge between $u$ and $v$ before. We get $G'$ in
which every vertex has even degree. So $G'$ must have an Eulerian circuit, call it $\sigma$.
It contains the edge $e$, but we can remove $e$ to get an Eulerian trail in the original graph $G$.\slug

Since the graph representing the K\"onigsberg bridge problem has four vertices, all of odd degree,
the two theorems we just proved tell us that it contains neither an Eulerian trail nor an
Eulerian circuit. Hence it is impossible to cross every bridge in K\"onigsberg exactly once.

\advsect Planar graphs

\datestamp{14}{xi} A graph $G$ is called {\it planar} if we can draw $G$ in the plane wihout edges
crossing. If we can, such a drawing is called a {\it planar embedding} of $G$.
The graph $K_4$ is planar; its representation in Fig.~2 does not make this evident, but
a planar embedding is given in Fig.~8. Likewise, the hypercube $Q_3$ was drawn with edge crossings
in Fig.~3, but it is also planar.
\midinsert
$$\epsfbox{planarembeddings.ps}$$
\vskip5pt
\caption{Planar embeddings of $K_4$ and $Q_3$, with numbered faces}
\endinsert
\goodbreak
Planarity is a property that passes down to subgraphs, as the following proposition shows.

\edef\propsubgraphplanar{\the\thmcount}
\proclaim Proposition \advthm. If $G$ is planar then any subgraph $H$ of $G$ is also planar.

\proof If $G$ is drawn in the plane without edges crossing, and we erase some edges and vertices
to create a drawing of $H$, we cannot introduce any edge crossings along the way.\slug

A planar embedding of a graph $G$ determines regions, or {\it faces} bounded by the edges,
including one region ``outside'' the graph.
We define the {\it Euler characteristic}
of a planar embedding of $G = (V,E)$ to be the quantity
$$\chi = |V| - |E| + f,$$
where $f$ is the number of faces in the planar embedding.
For the examples in Fig.~8, we have
$$\chi = 4 - 6 + 4 = 2$$
in the case of $K_4$, and
$$\chi = 8 - 12 + 6 = 2$$
in the case of $Q_3$. This might seem like a funny coincidence. The following theorem shows that
it is not.

\edef\thmeulersformula{\the\thmcount}
\parenproclaim Theorem {\advthm} (Euler's formula). Let $G = (V,E)$ be a connected planar graph,
let $f$ be the number of faces in a planar embedding of $G$. Then the Euler
characteristic $\chi$ of $G$ equals $2$.

The proof of this formula requires a theorem whose statement seems so ridiculously intuitive
that it seems silly that it even needs to be stated as a theorem. Despite its innocuous
appearance, is surprisingly difficult to prove, so much so that its original 1887 proof by
C.~Jordan was thought by many mathematicians to be so lacking in detail as to be wrong, and
the first widely accepted proof did not appear until 1905 (due to O.~Veblen). More recently
some mathematicians have analysed Jordan's original proof and found that it was more or less
correct to begin with, just not presented in a satisfactory way. In any case, here is the theorem.

\parenproclaim Theorem {\advthm} (Jordan curve theorem). Every closed curve in the plane
$\RR^2$ that does not intersect itself divides the plane into two distinct regions.\slug

The two regions that the closed curve creates are simply the inside of the curve and the outside
of the curve. Using this theorem, we can now prove Euler's formula.

\medskip\noindent{\it Proof of Theorem~\thmeulersformula}.\enspace
We proceed by induction on $m = |E|$. In the case that $m=0$, $G$ can only be $\bullet$,
in which case we have $\chi = 1 - 0 + 1 = 2$.

Now let $m\ge 0$ and assume that the theorem holds for all natural numbers at most $m$.
Let $G = (V,E)$ be a connected planar graph with $|E| = m+1$. Draw $G$ in the plane and let
$f$ be the number of faces in the embedding. There are two cases, according to whether
$G$ is a tree or not.
If $G$ is a tree, then $f=1$, since $G$ has no cycles. But we also have $|E| = |V|-1$, so
$$\chi = |V|-|E|+f = |V|-\bigl(|V|-1\bigr) + 1 = 2.$$
If $G$ is not a tree, then $G$ has a cycle, call it $\sigma$. Form $G'$ by removing
exactly one edge $e$ of $\sigma$; so
$$ |E'| = |E| - 1 = m+1-1 = m.$$
By the detour lemma, $G'$ is still connected, and it is planar, since removing an edge doesn't
create any crossings. By the inductive hypothesis, the Euler characteristic $\chi'$ of $G'$
satisfies
$$\chi' = |V| - |E'| + f' = 2,$$
where $f'$ is the number of faces in the drawing of $G$. Returning to the drawing
of $G$, the cycle $\sigma$ is a closed curve without intersections in the plane, so
it divides the plane into two distinct regions (each of which may have multiple faces within them).
In particular, the removed edge $e$ forms a border between two distinct faces, and in $G'$,
these two faces combine into one face. Hence $f = f'+1$. Putting everything together,
we calculate that the Euler characteristic $\chi$ of $G$ satisfies
$$\eqalign{
\chi &= |V| - |E| + f \cr
&= |V| - \bigl(|E'| + 1\bigr) + (f' + 1) \cr
&= |V| - |E'| + f' \cr
&= 2.\noskipslug\cr
}$$

Euler's formula tells us that the number of faces in any drawing of $G$ depends entirely on
the number of vertices and edges that $G$ has. We can use this information to give conditions
for a graph to be non-planar.

\edef\thmnonplanarconditions{\the\thmcount}
\proclaim Theorem \advthm. Let $G = (V,E)$ be a connected planar graph with $|V|\ge 5$. Then
$$|E| \le 3|V| - 6.$$
Under the further assumption that $G$ contains no triangles, we have the better bound
$$|E| \le 2|V| - 4.$$

\proof The proof is by double counting. Let $R$ be the set of all regions into which $G$
divides the plane, so that $|R| = f$, the number of faces. Consider the set
$$S = \bigl\{ (e, r) \subseteq E\times R : \hbox{the edge}\ e\ \hbox{touches the region}\ r\bigr\}.$$
We will count $|S|$ in two ways. First off, each edge $e\in E$ touches at most two regions,
so $|S|\le 2|E|$. On the other hand, every region is bounded by a cycle, and cycles
have at least $3$ edges, so each region $r\in R$ touches at least $3$ edges. In other words,
$|S|\ge 3f$. Chaining these two inequalities, we have
$$3f \le |S| \le 2|E|.$$
But by Euler's formula, we have $|V| - |E| + f = 2$, so $f = 2-|V|+|E|$, and substituting this above,
we get
$$3\bigl( 2-|V|+|E|\bigr) \le 2|E|.$$
Distributing and rearranging terms yields the desired inequality $|E| \le 3|V|-6$.

If we have the further assumption that there are no triangles in $G$, then every region must
touch at least $4$ edges, allowing us to conclude the stronger bound $4f \le 2|E|$. Then we proceed
as above to get
$$4\bigl( 2-|V|+|E|\bigr) \le 2|E|,$$
which can be manipulated to give $|E|\le 2|V|-4$.\slug

The condition that $|V|\ge 5$ is a complete non-issue, since every graph with fewer than $5$
vertices is a subgraph of $K_4$, and we can apply Proposition~{\propsubgraphplanar},
since we already know that $K_4$ is planar. As a corollary of the (contrapositive of the)
previous theorem, we can give two important examples of nonplanar graphs.

\proclaim Corollary \advthm.  The complete graph $K_5$ and the complete bipartite graph $K_{3,3}$
are nonplanar.

\proof If $(V,E) = K_5$ we have $|V| = 5$ so $3|V| - 6 = 9$,
but $|E| = 5\cdot 4 / 2 = 10$. So by the contrapositive of the previous theorem $K_5$ cannot
be planar.

The graph $K_{3,3}$ does not contain any triangles, so if $K_{3,3}$ is planar, then the
stronger bound $|V| \le 2|V|-4$ must hold. But in $(V,E) = K_{3,3}$, we have $|V| = 6$,
so $2|V|-4 = 8$, and we have $|E| = 9$. So $K_{3,3}$ is also nonplanar.\slug

\midinsert
$$\epsfbox{k5k33.ps}$$
\vskip5pt
\caption{The nonplanar graphs $K_5$ and $K_{3,3}$.}
\endinsert
\goodbreak

This corollary, combined with the contrapositive of Proposition~{\propsubgraphplanar},
shows that $K_n$ is nonplanar for all $n\ge 5$, and $K_{m,n}$ is non-planar whenever
$m$ and $n$ are both at least $3$. More generally, any graph that contains $K_5$ or $K_{3,3}$
as a subgraph is nonplanar.

However, subgraphs are not exactly the right notion to be considering
when talking about planarity. To see why, consider the graph in Fig.~10.
\midinsert
$$\epsfbox{k5dot.ps}$$
\vskip5pt
\caption{The graph $K_5$, with an extra vertex subdividing an edge.}
\endinsert
\goodbreak
It is nonplanar, since if it had a planar embedding, then by contracting one of the
edges incident on the extra vertex and bending other edges accordingly,
we would obtain a planar embedding of $K_5$.
On the other hand, $K_5$ is not a subgraph of this graph.
We thus introduce the following notion.
We say that a graph $H$ is a {\it graph minor} of a graph $G$ if $H$ can be obtained
from $G$ by repeatedly either
\medskip
\item{i)} deleting an edge;
\smallskip
\item{ii)} deleting a vertex; or
\smallskip
\item{iii)} contracting an edge $uv$ by removing it and merging $u$ and $v$ into a single vertex
(and also combining any resulting multiple edges into a single edge).
\medskip
Any number of the first two operations simply creates a subgraph. It is the contraction operation
that produces interesting examples of graph minors. It is easy to see pictorially that
contracting edges preserves planarity (we will not prove this rigorously), so {\sl any graph
minor of a planar graph is planar}. We can use this to show, for instance, that the Petersen graph,
depicted in Fig.~11, is nonplanar, since by contracting each of the five edges connecting
the inner star pentagram to the outer pentagon, we obtain $K_5$. (Note that
Theorem~{\thmnonplanarconditions} does not directly apply to the Petersen graph.)
\midinsert
$$\epsfbox{petersen.ps}$$
\vskip5pt
\caption{The Petersen graph.}
\endinsert
\goodbreak
\bye

\noindent\datestamp{22}{} \hskip\parindent
Not only are all graphs that have $K_5$ or $K_{3,3}$ as a minor nonplanar, it turns out
that these are the only nonplanar graphs!

\parenproclaim Theorem {\advthm} (Wagner's theorem). A graph $G$ is nonplanar if and only
if either $K_5$ or $K_{3,3}$ is a graph minor of $G$.\slug

This theorem was proven in 1937 by K.~Wagner, but it is closely related to a theorem that was
proved earlier (in 1930) by K.~Kuratowski.
It deals with the notion of a {\it subdivision} of a graph $G$:
this is any graph obtained by repeatedly placing vertices on edges of $G$, subdividing an edge
into two edges each time. (Fig.~10 was an example of a subdivision of $K_5$.) We now state Kuratowski's
theorem.

\parenproclaim Theorem {\advthm} (Kuratowski's theorem). A graph $G$ is nonplanar if and only
if it contains as a subgraph a subdivision of either $K_5$ or $K_{3,3}$.\slug

It is easy to
see that if $H$ is a subdivision of $G$, then $G$ is a minor of $H$, since we can reobtain
$G$ by contracting all the edges created by the subdivision operations. So the two theorems
above are certainly very closely related, though it is not immediately obvious if one should imply
the other. The truth is that they are equivalent, because it can be shown that any graph with
either $K_5$ or $K_{3,3}$ as a minor also has a subgraph that is a subdivision of one of them.

\vfill\eject
\begingroup\headline{\hfil}\footline{\hfil}
\vglue160pt
\centerline{\titlefont IV. COMBINATORICS}
\vskip220pt
\bigskip
\begingroup\obeylines\eightssi
\hfill Quare merito suo utilissima censenda est Ars, {\eightss Combinatoria} dicta,
\hfill quae huic mentis nostrae defectui medetur,
\hfill docetque sic enumerare modos omnes possibiles,
\hfill secundum quos res plures permisceri, transponi vel conjungi invicem possunt,
\hfill ut certi, simus, nos nullum eorum praetermisisse,
\hfill qui instituto nostro conducere valent.
\eightss
\smallskip
\hfill --- JAKOB BERNOULLI, {\eightssi Ars Conjectandi} (1713)
\endgroup%\obeylines
\bigskip\goodbreak
\vfill\eject

\advsect Basic counting techniques

Combinatorics is a branch of mathematics that, broadly speaking, deals with counting (or estimating)
the cardinalities of finite sets. Let us now recollect the counting techniques that we know.
If two sets $A$ and $B$ are disjoint, i.e., $A\cap B = \emptyset$,
then the number of elements in the union $A\cup B$ is simply $|A| + |B|$.
The cardinality of the power set $A\times B$ is $|A|\cdot |B|$, since for each of the $|A|$ elements
$a\in A$, there are $|B|$ choices for $b\in B$ to pair with it.
Lastly, if $U$ is a universe and $A\subseteq U$, then $\bar A$ is the set of all elements
of the universe that don't belong to $A$, so $|\bar A| = |U|-|A|$.

\medskip\boldlabel Counting strings.
Back in the proof of Proposition~{\propcardpowerset}, we counted the number of binary strings of
length $n$ to be $2^n$. More generally, if there are $s$ different symbols and $t$ different
slots in which to put them, then the number of strings is $s^t$. This can be phrased
in terms of the number of functions from a set $T$ of size $t$ to a set $S$ of size $s$.
Indeed, people often use the notation $S^T$ to denote the set of all functions from $T$ to $S$,
and the notation should remind you that $|S^T| = |S|^{|T|}$.

For example, using the symbol set $S = \{a,b,c,d\}$, the number of strings of length $5$
is $|S|^5 = 4^5$. What about the number of strings of length {\it at most $5$}, including
one empty string of length $0$? This is the number of strings of length $0$, plus the number
of strings of length $1$, and so on up to and including length $5$; that is, this equals
$$4^0 + 4^1 + 4^2 + 4^3 + 4^4 + 4^5.$$

@@
